maze-dataset

### 2025-03-30 docs/paper/paper.tex
wip

  \begin{itemize}
  \tightlist
  \item
    we also include the \texttt{RasterizedMazeDataset} class, utilizing
    \texttt{as\_pixels()}, in our codebase, which can exactly mimic the
    outputs provided in
    \texttt{easy-to-hard-data}(\protect\hyperlink{ref-easy_to_hard}{Schwarzschild,
    Borgnia, Gupta, Bansal, et al., 2021}) and can be configured to be
    similar to the outputs of
    (\protect\hyperlink{ref-gh_Nemeth_2019}{Németh, 2019}).
  \item
    Our \texttt{as\_ascii()} method provides a format similar to that
    used in Oppenheim
    (\protect\hyperlink{ref-gh-oppenheimj2018maze}{2018}).
  \item
    Our \texttt{MazePlot} class provides a feature-rich plotting utility
    with support for multiple paths, heatmaps over positions, and more.
    This is similar to the outputs of Ehsan
    (\protect\hyperlink{ref-gh_Ehsan_2022}{2022})
  \end{itemize}
\item
  The text format provided by \texttt{SolvedMaze(...).as\_tokens()} is
  similar to that of (\protect\hyperlink{ref-eval-LLM-graphs}{Liu \& Wu,
  2023}), but provides over 5.8 million unique formats for converting
  mazes to a text stream. We maintain a single underlying format,
  meaning that the same maze can be turned into a variety of text
  streams to assess how the precise format of the text stream affects
  the model.
\item
  For rigorous investigations of the response of a model to various
  distributional shifts, preserving metadata about the generation
  algorithm with the dataset itself is essential. To this end, our
  package efficiently stores the dataset along with its metadata in a
  single human-readable file (\protect\hyperlink{ref-zanj}{M.
  Ivanitskiy, n.d.}). This metadata is loaded when the dataset is
  retrieved from disk and makes it simple to understand how exactly each
  maze was generated. As far as we are aware, no existing packages do
  this reliably.
\item
  Storing mazes as images is not only difficult to work with, but also
  inefficient. Directly storing adjacency matrices is also inefficient
  as subgraphs of the lattice are sparse. Storing adjacency lists can be
  efficient, but comes with a higher lookup cost and possible high
  comparison cost. We use a simple, efficient representation of mazes
  that is optimized for subgraphs of a \(d\)-dimensional finite lattice
  that we do not believe is used in any existing maze generation
  package.
\item
  Our package is easily installable with source code freely available.
  It is extensively tested, type hinted, benchmarked, and documented.
  Many other maze generation packages lack this level of rigor and
  scope, and some (\protect\hyperlink{ref-ayaz2008maze}{Ayaz et al.,
  2008}) appear to simply no longer be accessible.
\end{itemize}

@@

\hypertarget{visual-output-formats}{%
\section{Visual Output Formats}\label{visual-output-formats}}

@@

Optional delimiter tokens may be added in many places in the output.
Delimiter options are all configured using the parameters named
\texttt{pre}, \texttt{intra}, and \texttt{post} in various
\texttt{\_TokenizerElement} classes. Each option controls a unique
delimiter token. Here we describe each \texttt{\_TokenizerElement} and
the behaviors they support. We also discuss some of the model behaviors
and properties that may be investigated using these options.

\hypertarget{coordtokenizer}{%
\subsubsection{Coordinates}\label{coordtokenizer}}

The \texttt{\_CoordTokenizer} object controls how coordinates in the
lattice are represented in across all token regions. Options include:

\begin{itemize}
\tightlist
\item
  \textbf{Unique tokens}: Each coordinate is represented as a single
  unique token \texttt{"(i,j)"}
\item
  \textbf{Coordinate tuple tokens}: Each coordinate is represented as a
  sequence of 2 tokens, respectively encoding the row and column
  positions: \texttt{{[}"i",\ ",",\ "j"{]}}
\end{itemize}

\hypertarget{adjlisttokenizer}{%
\subsubsection{Adjacency List}\label{adjlisttokenizer}}

The \texttt{\_AdjListTokenizer} object controls this token region. All
tokenizations represent the maze connectivity as a sequence of
connections or walls between pairs of adjacent coordinates in the
lattice.

\begin{itemize}
\tightlist
\item
  \texttt{\_EdgeSubset}: Specifies the subset of lattice edges to be
  tokenized

  \begin{itemize}
  \tightlist
  \item
    \textbf{All edges}: Every edge in the lattice
  \item
    \textbf{Connections}: Only edges which contain a connection
  \item
    \textbf{Walls}: Only edges which contain a wall
  \end{itemize}
\item
  \texttt{\_EdgePermuter}: Specifies how to sequence the two coordinates
  in each lattice edge

  \begin{itemize}
  \tightlist
  \item
    \textbf{Random}
  \item
    \textbf{Sorted}: The smaller coordinate always comes first
  \item
    \textbf{Both permutations}: Each edge is represented twice, once
    with each permutation. This option attempts to represent connections
    in a more directionally symmetric manner. Including only one
    permutation of each edge may affect models' internal representations
    of edges, treating a path traversing the edge differently depending
    on if the coordinate sequence in the path matches the sequence in
    the adjacency list.
  \end{itemize}
\item
  \texttt{shuffle\_d0}: Whether to shuffle the edges randomly or sort
  them in the output by their first coordinate
\item
  \texttt{connection\_token\_ordinal}: Location in the sequence of the
  token representing whether the edge is a connection or a wall
\end{itemize}

\hypertarget{pathtokenizer}{%
\subsubsection{Path}\label{pathtokenizer}}

The \texttt{\_PathTokenizer} object controls this token region. Paths
are all represented as a sequence of steps moving from the start to the
end position.

\begin{itemize}
\tightlist
\item
  \texttt{\_StepSize}: Specifies the size of each step

  \begin{itemize}
  \tightlist
  \item
    \textbf{Singles}: Every coordinate traversed between start and end
    is directly represented
  \item
    \textbf{Forks}: Only coordinates at forking points in the maze are
    represented. The paths between forking points are implicit. Using
    this option might train models more directly to represent forking
    points differently from coordinates where the maze connectivity
    implies an obvious next step in the path.
  \end{itemize}
\item
  \texttt{\_StepTokenizer}: Specifies how an individual step is
  represented

  \begin{itemize}
  \tightlist
  \item
    \textbf{Coordinate}: The coordinates of each step are directly
    tokenized using a \texttt{\_CoordTokenizer}
  \item
    \textbf{Cardinal direction}: A single token corresponding to the
    cardinal direction taken at the starting position of that step.
    E.g., \texttt{NORTH}, \texttt{SOUTH}. If using a \texttt{\_StepSize}
    other than \textbf{Singles}, this direction may not correspond to
    the final direction traveled to arrive at the end position of the
    step.
  \item
    \textbf{Relative direction}: A single token corresponding to the
    first-person perspective relative direction taken at the starting
    position of that step. E.g., \texttt{RIGHT}, \texttt{LEFT}.
  \item
    \textbf{Distance}: A single token corresponding to the number of
    coordinate positions traversed in that step. E.g., using a
    \texttt{\_StepSize} of \textbf{Singles}, the \textbf{Distance} token
    would be the same for each step, corresponding to a distance of 1
    coordinate. This option is only of interest in combination with a
    \texttt{\_StepSize} other than \textbf{Singles}.
  \end{itemize}
\end{itemize}

A \texttt{\_PathTokenizer} contains a sequence of one or more unique
\texttt{\_StepTokenizer} objects. Different step representations may be
mixed and permuted, allowing for investigation of model representations
of multiple aspects of a maze solution at once.

\hypertarget{token-training}{%
\subsection{Tokenized Outputs for Training and
Evaluation}\label{token-training}}

During deployment we provide only the prompt up to the
\texttt{\textless{}PATH\_START\textgreater{}} token.

Examples of usage of this dataset to train autoregressive transformers
can be found in our \texttt{maze-transformer} library
(\protect\hyperlink{ref-maze-transformer-github}{Michael I. Ivanitskiy
et al., 2023b}). Other tokenization and vocabulary schemes are also
included, such as representing each coordinate as a pair of \(i,j\)
index tokens.

\hypertarget{extensibility}{%
\subsection{Extensibility}\label{extensibility}}

@@

\hypertarget{benchmarks}{%
\section{Benchmarks of Generation Speed}\label{benchmarks}}

@@

@@

This package utilizes a simple, efficient representation of mazes. Using
an adjacency list to represent mazes would lead to a poor lookup time of
whether any given connection exists, whilst using a dense adjacency
matrix would waste memory by failing to exploit the structure (e.g.,
only 4 of the diagonals would be filled in). Instead, we describe mazes
with the following simple representation: for a \(d\)-dimensional
lattice with \(r\) rows and \(c\) columns, we initialize a boolean array
\(A = \{0, 1\}^{d \times r \times c}\), which we refer to in the code as
a \texttt{connection\_list}. The value at \(A[0,i,j]\) determines
whether a downward connection exists from node \([i,j]\) to
\([i+1, j]\). Likewise, the value at \(A[1,i,j]\) determines whether a
rightwards connection to \([i, j+1]\) exists. Thus, we avoid duplication
of data about the existence of connections, at the cost of requiring
additional care with indexing when looking for a connection upwards or
to the left. Note that this setup allows for a periodic lattice.

To produce solutions to mazes, two points are selected uniformly at
random without replacement from the connected component of the maze, and
the \(A^*\) algorithm (\protect\hyperlink{ref-A_star}{Hart et al.,
1968}) is applied to find the shortest path between them.

Parallelization is implemented via the \texttt{multiprocessing} module
in the Python standard library, and parallel generation can be
controlled via keyword arguments to the
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#MazeDataset.from_config}{\texttt{MazeDataset.from\_config()}}
function.

\hypertarget{limitations}{%
\subsection{\texorpdfstring{Limitations of
\texttt{maze-dataset}}{Limitations of maze-dataset}}\label{limitations}}

For simplicity, the package primarily supports mazes that are sub-graphs
of a 2-dimensional rectangular lattice. Some support for
higher-dimensional lattices is present, but not all output formats are
adapted for higher dimensional mazes.
\protect\hyperlink{implementation-implementation}{Implementation}
\protect\hyperlink{implementation-implementation}{Implementation}

@@

>>>>

  \begin{itemize}
  \tightlist
  \item
    we also include the \texttt{RasterizedMazeDataset} class, utilizing
    \texttt{as\_pixels()}, in our codebase, which can exactly mimic the
    outputs provided in
    \texttt{easy-to-hard-data}(\protect\hyperlink{ref-easy_to_hard}{Schwarzschild,
    Borgnia, Gupta, Bansal, et al., 2021}) and can be configured to be
    similar to the outputs of
    (\protect\hyperlink{ref-gh_Nemeth_2019}{Németh, 2019}).
  \item
    Our \texttt{as\_ascii()} method provides a format similar to that
    used in Oppenheim
    (\protect\hyperlink{ref-gh-oppenheimj2018maze}{2018}).
  \item
    Our \texttt{MazePlot} class provides a feature-rich plotting utility
    with support for multiple paths, heatmaps over positions, and more.
    This is similar to the outputs of Ehsan
    (\protect\hyperlink{ref-gh_Ehsan_2022}{2022})
  \end{itemize}
\item
  The text format provided by \texttt{SolvedMaze(...).as\_tokens()} is
  similar to that of (\protect\hyperlink{ref-eval-LLM-graphs}{Liu \& Wu,
  2023}), but provides over 5.8 million unique formats for converting
  mazes to a text stream. We maintain a single underlying format,
  meaning that the same maze can be turned into a variety of text
  streams to assess how the precise format of the text stream affects
  the model.
\item
  For rigorous investigations of the response of a model to various
  distributional shifts, preserving metadata about the generation
  algorithm with the dataset itself is essential. To this end, our
  package efficiently stores the dataset along with its metadata in a
  single human-readable file (\protect\hyperlink{ref-zanj}{M.
  Ivanitskiy, n.d.}). This metadata is loaded when the dataset is
  retrieved from disk and makes it simple to understand how exactly each
  maze was generated. As far as we are aware, no existing packages do
  this reliably.
\item
  Storing mazes as images is not only difficult to work with, but also
  inefficient. Directly storing adjacency matrices is also inefficient
  as subgraphs of the lattice are sparse. Storing adjacency lists can be
  efficient, but comes with a higher lookup cost and possible high
  comparison cost. We use a simple, efficient representation of mazes
  that is optimized for subgraphs of a \(d\)-dimensional finite lattice
  that we do not believe is used in any existing maze generation
  package. This method is described in detail in
  \autoref{implementation}.
\item
  Our package is easily installable with source code freely available.
  It is extensively tested, type hinted, benchmarked, and documented.
  Many other maze generation packages lack this level of rigor and
  scope, and some (\protect\hyperlink{ref-ayaz2008maze}{Ayaz et al.,
  2008}) appear to simply no longer be accessible.
\end{itemize}

@@

\hypertarget{visual-output-formats}{%
\subsection{Visual Output Formats}\label{visual-output-formats}}

@@

\textbf{TODO: add fst stuff}

@@

\hypertarget{benchmarks}{%
\subsection{Benchmarks of Generation Speed}\label{benchmarks}}

@@

\hypertarget{success-rate-estimation}{%
\subsection{Success Rate Estimation}\label{success-rate-estimation}}

In order to replicate the exact dataset distribution of
(\protect\hyperlink{ref-easy_to_hard}{Schwarzschild, Borgnia, Gupta,
Bansal, et al., 2021}), we allow placing additional constraints in
\texttt{MazeDatasetConfig.endpoint\_kwargs:\ EndpointKwargsType}, such
as enforcing that the start or end point be in a ``dead end'' with only
one accessible neighbor cell. However, combining this with cyclic mazes
(such as those generated with percolation), as was required for the work
in (\protect\hyperlink{ref-knutson2024logicalextrapolation}{Knutson et
al., 2024}), can lead to an absence of valid start and end points.
Placing theoretical bounds on this success rate is difficult, as it
depends on the exact maze generation algorithm and parameters used.
However, our package provides a way to estimate the success rate of a
given configuration using a symbolic regression model trained with PySR
(\protect\hyperlink{ref-pysr}{Cranmer, 2023}). More details on this can
be found in
\href{https://understanding-search.github.io/maze-dataset/notebooks/estimate_dataset_fractions.html}{\texttt{estimate\_dataset\_fractions.ipynb}}.

Using the estimation simply requires the user to call
\texttt{cfg\_new:\ MazeDatasetConfig\ =\ cfg.success\_fraction\_compensate()}
given their predefined \texttt{cfg}.

\hypertarget{success-rate-estimation}{%
\subsubsection{Success Rate Estimation
Algorithm}\label{success-rate-estimation}}

The base function learned by symbolic regression is not particularly
insightful, and is potentially subject to change. It is defined as
\href{https://understanding-search.github.io/maze-dataset/maze_dataset/dataset/success_predict_math.html\#cfg_success_predict_fn}{\texttt{cfg\_success\_predict\_fn}},
and takes a 5 dimensional float vector created by
\texttt{MazeDatasetConfig.\_to\_ps\_array()} which represents the 0)
percolation value 1) grid size 2) endpoint deadend configuration 3)
endpoint uniqueness 4) categorical generation function index.

However, the outputs of this function are not directly usable due to
minor divergences at the endpoints with respect to the percolation
probability \(p\). Since we know that maze success is either guaranteed
or impossible for \(p=0\) and \(p=1\), we define the \texttt{soft\_step}
function to nudge the raw output of the symbolic regression. This
function is defined with:

\begin{itemize}
\item
  \textbf{Shifted Sigmoid}: Creates sharp transition at \(x=0.5\)
  \[\sigma_s(x) = (1 + e^{-10^3 \cdot (x-0.5)})^{-1}\]
\item
  \textbf{h-function}: Transition function combining polynomial and
  sigmoid components
  \[h(q,a) = q \cdot (1 - |2q-1|^a) \cdot (1-\sigma_s(q)) + (1-(1-q) \cdot (1 - |2(1-q)-1|^a)) \cdot \sigma_s(q)\]
\item
  \textbf{Amplitude Scaling}: Weight-modulated polynomial
  \[A(q,a,w) = w \cdot (1 - |2q-1|^a)\]
\item
  \textbf{Soft Step}: Identity-like for \(p \approx 0.5\), pushes \(x\)
  to extremes otherwise
  \[\text{soft\_step}(x, p, \alpha, w) = h(x, A(p, \alpha, w))\]
\end{itemize}

Finally, we define

\[
  \text{cfg\_success\_predict\_fn}(\mathbf{x}) = \text{soft\_step}(\text{raw\_val}, x_0, 5, 10)
\]

where \texttt{raw\_val} is the output of the symbolic regression model.
\(x_0\) is the percolation probability, while all other parameters from
\texttt{\_to\_ps\_array()} only affect \texttt{raw\_val}.

\begin{figure}
\centering
\includegraphics{figures/ep/ep_deadends_unique.pdf}
\caption{An example of both empirical and predicted success rates as a
function of the percolation probability \(p\) for various maze sizes,
percolation with and without depth first search, and
\texttt{endpoint\_kwargs} requiring that both the start and end be in
unique dead ends.}
\end{figure}

@@

This package utilizes a simple, efficient representation of mazes. Using
an adjacency list to represent mazes would lead to a poor lookup time of
whether any given connection exists, while using an adjacency matrix
would waste memory by failing to exploit the structure (e.g., only 4 of
the diagonals would be filled in). Instead, we describe mazes with the
following simple representation: for a \(d\)-dimensional lattice with
\(r\) rows and \(c\) columns, we initialize a boolean array
\(A = \{0, 1\}^{d \times r \times c}\), which we refer to in the code as
a \texttt{connection\_list}. The value at \(A[0,i,j]\) determines
whether a downward connection exists from node \([i,j]\) to
\([i+1, j]\). Likewise, the value at \(A[1,i,j]\) determines whether a
rightwards connection to \([i, j+1]\) exists. Thus, we avoid duplication
of data about the existence of connections, at the cost of requiring
additional care with indexing when looking for a connection upwards or
to the left. Note that this setup allows for a periodic lattice.

To produce solutions to mazes, two points are selected uniformly at
random without replacement from the connected component of the maze, and
the \(A^*\) algorithm (\protect\hyperlink{ref-A_star}{Hart et al.,
1968}) is applied to find the shortest path between them. The endpoint
selection can be affected by
\texttt{MazeDatasetConfig.endpoint\_kwargs:\ EndpointKwargsType}, and
complications caused by this are detailed in
\autoref{success-rate-estimation}.

Parallelization is implemented via the \texttt{multiprocessing} module
in the Python standard library, and parallel generation can be
controlled via keyword arguments to
\href{https://understanding-search.github.io/maze-dataset/maze_dataset.html\#MazeDataset.from_config}{\texttt{MazeDataset.from\_config()}}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-pysr}{}}%
Cranmer, M. (2023). Interpretable machine learning for science with PySR
and SymbolicRegression. jl. \emph{arXiv Preprint arXiv:2305.01582}.
